### 2015-09-22 Manuscript/SW_early_life_telomeres_v9.Rmd
More updates

#####Early-life telomere length and age
Mean ($\pm$ standard error) telomere length in first-year birds was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb. This is lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb). However, when we considered age classes within the first year of life separately, we found substantial age-related variation in telomere length (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~Ageclass,data = juv))[1,5])`). Chicks had longer telomeres than any other age group, but there was an apparent increase in average telomere length from the fledgling to subadult stages (Fig. 1A). However, a longitudinal analysis of telomere loss showed that chicks, fledglings and subadults all lost telomeres early in life, but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig. 1B). This suggests that the apparent increase in telomere length between fledglings and subadults in the cross-sectional data was a result of the sampling, rather than telomere elongation.

@@

The rate of telomere shortening in early life did not vary significantly among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`); however, what variation there was in rates of telomere shortening could be explained by temporal variation in food availability (R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`). This pattern corresponded with the cross-sectional data: individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (Fig. 2B).  Unfortunately our limited sample size within the longitudinal dataset precluded us from running this analysis separately for each age class.

@@

For the subset of individuals with longitudinal data, we tested whether the factors previously identified as being related to early-life telomere length (i.e. tarsus length, sex and number of helpers) were also associated with differences in rates of early life telomere loss. Due to a limited sample size, we considered all age classes together for this analysis. We found that none of these variables were associated with differences in telomere loss (all *P* > 0.05).

#####Early life telomere dynamics and survival
At the population-level, we found no evidence that individuals born in breeding seasons with high food availability had longer lifespans (R^2^ = `r round2(summary(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,5])`), nor that individuals born in years where average telomere length was longer had longer lifespans (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig. 5A). Testing the latter relationship separately for each age class revealed that there was no age-specific relationship between cohort-level telomere length and lifespan (all *P* > 0.2). There was a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig. 5B).

For the individual-based survival analysis, a log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`). A survival model including all first year birds, early-life telomere length was not associated with survival (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`). However, we did find differential survival effects among age classes. Chick telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; Fig. 6A). However, telomere length at both the fledgling and subadult stage was significantly related to survival (fledglings: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; subadults: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`). In both cases this effect was positive, with longer telomeres in early life associated with increased survival later in life (Figs 6B, 6C). Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).

>>>>

#####Early-life telomere length and age
Mean ($\pm$ standard error) telomere length in first-year birds was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb. This is lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb). However, when we considered age classes within the first year of life separately, we found substantial age-related variation in telomere length (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round3(anova(lm(TLKB~Ageclass,data = juv))[1,5])`). Chicks had longer telomeres than any other age group, but there was an apparent increase in average telomere length from the fledgling to subadult stages (Fig. 1A). However, a longitudinal analysis of telomere loss showed that chicks, fledglings and subadults all lost telomeres early in life, but that there was no difference in the rate of telomere loss among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig. 1B). This suggests that the apparent increase in telomere length between fledglings and subadults in the cross-sectional data was a result of the sampling and selective mortality, rather than telomere elongation.

@@

The rate of telomere shortening in early life (taken from the smaller subsert of birds with longitudinal samples) did not vary significantly among breeding seasons (F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`); however, what variation there was in rates of telomere shortening could be explained by temporal variation in food availability (R^2^ = `r round2(summary(lm(Insect~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Insect~TROCmean,data = loss.season,na.action=na.exclude))[1,5])`). This pattern corresponded with the cross-sectional data: individuals born in seasons with low food availability lost telomeres, on average, at a faster rate than those born in years with high food availability (Fig. 2B).  Unfortunately our limited sample size within the longitudinal dataset precluded us from running this analysis separately for each age class.

@@

For the subset of individuals with longitudinal data, we tested whether the factors previously identified as being related to early-life telomere length (i.e. tarsus length and number of helpers) were also associated with differences in rates of early life telomere loss. Due to a limited sample size, we considered all age classes together for this analysis. We found that none of these variables were associated with differences in telomere loss (all *P* > 0.05).

#####Early life telomere dynamics and survival
At the population-level, we found no evidence that individuals born in breeding seasons with high food availability had longer lifespans (R^2^ = `r round2(summary(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~Insect,data=juvseason,na.action=na.exclude))[1,5])`), nor that individuals born in years where average telomere length was longer had longer lifespans (R^2^ = `r round3(summary(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TLKBmean,data=juvseason,na.action=na.exclude))[1,5])`; Fig. 5A). Testing the latter relationship separately for each age class revealed that there was no age-specific relationship between cohort-level telomere length and lifespan (all *P* > 0.2). There was, however, a negative relationship between population-level telomere loss and lifespan, although this was marginally non-significant (R^2^ = `r round3(summary(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,4])`; *P* = `r round2(anova(lm(Lifespan~TROCmean,data=loss.season,na.action=na.exclude))[1,5])`; Fig. 5B).

For the individual-based survival analysis, a log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution (Weibull) = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`). A survival model including all first year birds showed that early-life telomere length was not associated with survival (estimate = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(juvall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`). However, we did find differential survival effects among age classes. Chick telomere length had no effect on survival  (estimate = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(chickall,Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; Fig. 6A), but telomere length at both the fledgling and subadult stage was significantly related to survival (fledglings: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'FL'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`; subadults: estimate = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,1], lessthan = F)`, CI = `r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(subset(FlSAall,Ageclass == 'SA'),Surv(Lifespan+1,Died)~cenTL), dist="lnorm")$res[3,3], lessthan = F)`). In both cases this effect was positive, with longer telomeres in early life associated with increased survival later in life (Figs 6B, 6C). Finally, using the longitudinal data, we found no effect of early-life telomere shortening on subsequent survival (estimate = `r round2(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,1])`, CI = `r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,2], lessthan = F)`-`r round3(flexsurvreg(with(Loss,Surv(Lifespan+1,Died)~cenTROC), dist="lnorm")$res[3,3], lessthan = F)`).
